Bibat: Batteries-include Bayesian Analysis Template

### 2024-02-02 example_projects/baseball/docs/report.qmd
copier

```{.python include="../baseball/data_preparation.py" start-line=74 end-line=159}
```

@@

```{.python include="../baseball/data_preparation.py" start-line=25 end-line=32}
```

@@

```{.python include="../baseball/stan_input_functions.py" start-line=11 end-line=34}
```

But why stop there? It can also be useful to generate Stan input data with a
model, by running it in simulation mode with hardcoded parameter values. Here
are some functions that do this for both of our models:
^[These functions require some more imports: `numpy` and `scipy.special.expit`]
```{.python include="../baseball/stan_input_functions.py" start-line=36}
```

>>>>

```{.python include="../baseball/data_preparation.py" start-line=75 end-line=159}
```

@@

```{.python include="../baseball/data_preparation.py" start-line=162 end-line=182}
```

@@

```{.python include="../baseball/stan_input_functions.py" start-line=22 end-line=46}
```

But why stop there? It can also be useful to generate Stan input data with a
model, by running it in simulation mode with hardcoded parameter values. Here
are some functions that do this for both of our models:
^[These functions require some more imports: `numpy` and `scipy.special.expit`]
```{.python include="../baseball/stan_input_functions.py" start-line=49}
```

### 2024-02-02 paper/main.qmd
copier

The idea that Bayesian statistical analysis comprises not just
inference, but also specific approaches to related activities like data
preparation, model design, diagnosis, debugging and criticism, dates
at least to @boxBayesianInferenceStatistical1992. There is increasing
scholarly recognition of the need for a holistic view of "Bayesian
workflow" [@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019]. Relatedly, software tools now
exist that address most individual aspects of a Bayesian workflow: see
@strumbeljPresentFutureSoftware for a review of the state of the art.

Unfortunately, currently available tools typically address one, or at most
a handful, of Bayesian workflow activities; it is left to the individual
project team to orchestrate the components. Writing software that performs
this orchestration can be time-consuming and tricky, especially in the common
scenario where it is not initially clear how many, or what kind of, statistical
models, datasets, data manipulations or investigations an analysis will require.

Bibat addresses this difficulty by providing a full-featured, high-quality
Bayesian workflow project that can be extended to implement a wide range of
statistical analyses.

@@

## Installation and usage

@@

## Documentation

>>>>

The term "Bayesian workflow" captures the idea that Bayesian statistical
analysis comprises not just inference, but also specific approaches to related
activities like data preparation, model design, diagnosis, debugging and
criticism. This idea can be found in @boxBayesianInferenceStatistical1992
and has recently received increasing scholarly recognition of
[@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019]. Software tools now exist
that address most individual aspects of a Bayesian workflow: see
@strumbeljPresentFutureSoftware for a review of the state of the art.

Unfortunately, each tool typically addresses one, or at most a few, of the many
activities that comprise a real Bayesian workflow software project; it is left
to the individual project team to orchestrate all of the tools they require.
Writing software that performs this orchestration can be time-consuming and
tricky, especially in the common scenario where it is not initially clear how
many, or what kind of, statistical models, datasets, data manipulations or
investigations an analysis will require.

Bibat is a new tool that addresses the difficulty of orchestrating Bayesian
workflow software by providing a full-featured, high-quality project that can be
extended to implement a wide range of statistical analyses.

@@

@@

### 2024-02-02 paper/main.tex
copier

\begin{document}
\maketitle
\begin{abstract}
Bayesian statistical workflow offers a powerful way to learn from data,
but software software projects that implement complex Bayesian workflows
in practice are unusual, partly due to the difficulty of orchestrating
Bayesian statistical software. Bibat addresses this challenge by
providing a full-featured, scalable Bayesian statistical analysis
project using an interactive template. Bibat is available on the Python
Package index, documented at \url{https://bibat.readthedocs.io/} and
developed at \url{https://github.com/teddygroves/bibat/}. This paper
explains the motivation for bibat, describes intended usage, discusses
bibat's design goals and how key challenges that were overcome in the
course of development, compares bibat with similar software, highlights
several examples of bibat's use in science and provides links to
community resources associated with bibat.
\end{abstract}

@@

The idea that Bayesian statistical analysis comprises not just
inference, but also specific approaches to related activities like data
preparation, model design, diagnosis, debugging and criticism, dates at
least to Box and Tiao (1992). There is increasing scholarly recognition
of the need for a holistic view of ``Bayesian workflow'' (Gelman et al.
2020; Grinsztajn et al. 2021; Gabry et al. 2019). Relatedly, software
tools now exist that address most individual aspects of a Bayesian
workflow: see Štrumbelj et al. (2024) for a review of the state of the
art.

Unfortunately, currently available tools typically address one, or at
most a handful, of Bayesian workflow activities; it is left to the
individual project team to orchestrate the components. Writing software
that performs this orchestration can be time-consuming and tricky,
especially in the common scenario where it is not initially clear how
many, or what kind of, statistical models, datasets, data manipulations
or investigations an analysis will require.

Bibat addresses this difficulty by providing a full-featured,
high-quality Bayesian workflow project that can be extended to implement
a wide range of statistical analyses.

@@

\subsection{Installation and usage}\label{installation-and-usage}

@@

\subsection{Documentation}\label{documentation}

@@

Bibat's design was informed by these considerations.

\subsection{Accommodate a wide range of statistical
analyses}\label{accommodate-a-wide-range-of-statistical-analyses}

Bibat projects explicitly allow for plurality at the level of input
datasets, prepared datasets, statistical models, fitting modes,
computation methods and analyses. In addition, bibat ensures that there
are minimal restrictions on the kind of components: for example,
datasets need not be singular or tabular, and statistical models need
not be representable in formula syntax.

Thanks to these accommodations a project team using bibat should
typically not need to foresee the ultimate requirements of their
analysis before starting the project.

\subsection{Encourage reproducibility}\label{encourage-reproducibility}

Bibat provides a preconfigured makefile with a target \texttt{analysis}
triggering creation of an isolated environment, installation of
dependencies, data preparation, statistical computation and analysis of
results. In this way a bibat analysis can be reproduced on most
platforms using a single command.

In particular, this target attempts to install cmdstan if necessary,
using a recipe tailored to the host operating system. This functionality
addresses a common issue where researchers find it difficult to install
Stan, especially on Windows.

Bibat also provides its Python code in the form of a package configured
using modern conventions for specifying dependencies and configuring
tooling, so that it is easy to maintain reproducibility as the analysis
develops.

\subsection{Use widely-adopted, open source and actively developed
tools}\label{use-widely-adopted-open-source-and-actively-developed-tools}

Bibat projects are is written in modern Python and uses pydantic
Pydantic developers (2022) and pandera (Niels Bantilan 2020) for data
modelling, Stan (Carpenter et al. 2017) for statistical inference,
cmdstanpy (Stan Development Team 2022) for python-Stan interface, arviz
(Kumar et al. 2019) for storing and analyzing inferences and sphinx
(Georg Brandl and the Sphinx team 2022) and quarto (Allaire et al. 2022)
for documentation.

Bibat itself is also written in modern Python and uses the popular tools
cookiecutter (Greenfeld et al. 2021), pydantic and click (Click
Developers 2022).

\subsection{Implement community standards and best practices for
collaborative software
development}\label{implement-community-standards-and-best-practices-for-collaborative-software-development}

Bibat projects include a preconfigured test environment, continuous
integration, linting and pre-commit hooks, making them suitable for
collaborative software development. In addition, including documentation
as a first class component of the analysis addresses a common problem in
academic statistics projects where the paper gets out of sync with the
code.

Bibat itself is continuously tested to ensure that it works on the
operating systems Linux, macOS and Windows. Bibat's continuous
integration runs a test suite as well as an end-to-end functional test
on all supported Python versions.

\section{Challenges}\label{challenges}

This section describes some specific challenges that often affect
Bayesian workflow projects and how bibat addresses them.

\subsection{Address complexity using a modular, file based
approach}\label{address-complexity-using-a-modular-file-based-approach}

As discussed in Gelman et al. (2020), Bayesian workflows are
complicated, featuring plurality, cyclicity and complexity at many
levels. Bibat accommodates this by separating non-interacting analysis
components into modules and by serialising data to files wherever
possible. Prepared datasets, statistical models, inference
configurations, inference results, plots and analyses all have file
representations. Fitting modes, data manipulations and data models are
modularised in code through the use of appropriately structured data
classes and functions.

Thanks to this approach it is possible to perform small sub analyses
individually and to iteratively add components without needing to
consider everything at once.

\subsection{Fitting modes}\label{fitting-modes}

As part of a Bayesian workflow it can be necessary to fit a model and
dataset in different ways. For example, one might perform MCMC sampling
of both the prior and posterior distributions, perform multiple
leave-out-one-fold fits for cross-validation or need to compare MCMC
sampling with an optimisation-based alternative.

Bibat accommodates this need by introducing an abstraction called
``fitting mode''. This abstraction allows bibat projects to handle
fitting a model and dataset in different ways appropriately and
flexibly. For example, the provided prior sampling fitting mode creates
a Stan input dictionary with the \texttt{likelihood} data variable set
to \texttt{0}, performs MCMC sampling and writes data to the
\texttt{InferenceData} group \texttt{prior}. Bibat provides fitting
modes corresponding for prior sampling, posterior sampling and k-fold
posterior sampling. Users can easily add additional fitting modes by
copying these examples.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Bayesian statistical workflow offers a powerful way to learn from data,
but software software projects that implement complex Bayesian workflows
in practice are unusual, partly due to the difficulty of orchestrating
Bayesian statistical software. Bibat addresses this challenge by
providing a full-featured, scalable Bayesian statistical analysis
project using an interactive template. Bibat is available on the Python
Package index, documented at \url{https://bibat.readthedocs.io/} and
developed at \url{https://github.com/teddygroves/bibat/}. This paper
explains the motivation for bibat, describes intended usage, discusses
bibat's design, compares bibat with similar software, highlights several
examples of bibat's use in science and provides links to community
resources associated with bibat.
\end{abstract}

@@

The term ``Bayesian workflow'' captures the idea that Bayesian
statistical analysis comprises not just inference, but also specific
approaches to related activities like data preparation, model design,
diagnosis, debugging and criticism. This idea can be found in Box and
Tiao (1992) and has recently received increasing scholarly recognition
of (Gelman et al. 2020; Grinsztajn et al. 2021; Gabry et al. 2019).
Software tools now exist that address most individual aspects of a
Bayesian workflow: see Štrumbelj et al. (2024) for a review of the state
of the art.

Unfortunately, each tool typically addresses one, or at most a few, of
the many activities that comprise a real Bayesian workflow software
project; it is left to the individual project team to orchestrate all of
the tools they require. Writing software that performs this
orchestration can be time-consuming and tricky, especially in the common
scenario where it is not initially clear how many, or what kind of,
statistical models, datasets, data manipulations or investigations an
analysis will require.

Bibat is a new tool that addresses the difficulty of orchestrating
Bayesian workflow software by providing a full-featured, high-quality
project that can be extended to implement a wide range of statistical
analyses.

@@

@@

@@

Bibat's design was informed by the aims to accommodate the many sources
of complexity and in a Bayesian workflow project, to ensure easy
reproducibility and to integrate many open-source, widely-adopted and
powerful Bayesian workflow tools and to encourage collaborative
development.

As discussed in Gelman et al. (2020), Bayesian workflows are
complicated, featuring plurality, cyclicity and complexity at many
levels. As a specialised Bayesian workflow template, a key goal for
bibat was to manage this complexity. Bibat achieves this aim by
separating non-interacting analysis components into separate,
potentially plural modules and by serialising data to files wherever
possible. Prepared datasets, statistical models, inference
configurations, inference results, plots and analyses all have file
representations. Fitting modes, data manipulations and data models are
modularised in code through the use of appropriately structured data
classes and functions. Thanks to this modular approach it is possible to
perform small sub analyses individually and to iteratively expand the
analysis by adding components without needing to consider everything at
once. In addition, bibat ensures that there are minimal restrictions on
the components: for example, datasets need not be singular or tabular
and statistical models need not be representable in formula syntax.
Thanks to these accommodations a project team using bibat should
typically not need to foresee the ultimate requirements of their
analysis before starting the project.

Bibat encourages reproducibility mainly by providing a preconfigured
makefile with a target \texttt{analysis} that triggers creation of an
isolated environment, installation of dependencies, data preparation,
statistical computation and analysis of results. In this way a bibat
analysis can be reproduced on most platforms using a single command. In
particular, this target attempts to install cmdstan if necessary, using
a recipe tailored to the host operating system. This functionality
addresses a common issue where researchers find it difficult to install
Stan, especially on Windows. A second way in which bibat encourages
reproducibility is by providing its Python code in the form of a
preconfigured package. Thanks to the use of modern conventions for
specifying dependencies and configuring tooling, we expect that this
will make bibat analyses easier for other researchers to work with,
thereby facilitating replication while also making initial development
more convenient.

Bibat integrates many widely-adopted, open-source and powerful tools to
implement the components of a Bayesian workflow. Bibat projects are is
written in modern Python and uses pydantic Pydantic developers (2022)
and pandera (Niels Bantilan 2020) for data modelling, Stan (Carpenter et
al. 2017) for statistical inference, cmdstanpy (Stan Development Team
2022) for python-Stan interface, arviz (Kumar et al. 2019) for storing
and analyzing inferences and sphinx (Georg Brandl and the Sphinx team
2022) and quarto (Allaire et al. 2022) for documentation. Bibat itself
is also written in modern Python and uses the popular tools cookiecutter
(Greenfeld et al. 2021), pydantic and click (Click Developers 2022).

To encourage collaborative development of Bayesian workflow projects,
Bibat projects include a preconfigured test environment, continuous
integration, linting and pre-commit hooks. In addition, including
documentation as a first class component of the analysis addresses a
common problem in academic statistics projects where the paper gets out
of sync with the code. Bibat itself is continuously tested to ensure
that it works on the operating systems Linux, macOS and Windows. Bibat's
continuous integration runs a test suite as well as an end-to-end
functional test on all supported Python versions.

\section{Fitting modes}\label{fitting-modes}

The most novel part of bibat's design is the introduction of an
abstraction called ``fitting mode'', which allows bibat project to
handle different ways of fitting a model to a dataset appropriately and
flexibly. This is often necessary as part of a Bayesian workflow: for
example, one might perform MCMC sampling of both the prior and posterior
distributions, perform multiple leave-out-one-fold fits for
cross-validation or need to compare MCMC sampling with an
optimisation-based alternative.

Fitting modes in bibat projects take the form of instances of the class
\texttt{FittingMode}. Each fitting mode contains a name, a function that
fits a prepared dataset and instructions for how and where to save the
results. For example, the provided prior sampling fitting mode creates a
Stan input dictionary with the \texttt{likelihood} data variable set to
\texttt{0}, performs MCMC sampling and writes data to the
\texttt{InferenceData} group \texttt{prior}. This design allows multiple
fits to be stored alongside each other naturally, facilitating many
common Bayesian workflow analyses, and can easily be configured to
prevent unnecessary computations.

Bibat provides fitting modes corresponding for prior sampling, posterior
sampling and k-fold posterior sampling. Users can easily add additional
fitting modes by copying these examples, or modify the
\texttt{FittingMode} class to achieve even richer functionality.
